Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-07 paper.tex
more stuff for text domain

How should we perform inference for the model
in~\ref{graphicalModel}(a)?
We take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a wake cycle means using the recognition model to search for programs that solve the tasks,
while a sleep cycle means updating the DSL and the recognition model
to more closely match the distribution of tasks.
This wake/sleep style of inference is the reason why we call our approach \system.

@@

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in each of these domains (Table~\ref{initialExampleDSL}).
 \begin{table}
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\texttt{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \texttt{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \texttt{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists& \texttt{(map (lambda (x) (+ x a)) l)}\\
   &\hspace{0.5cm} \emph{(add a to every element of list l)}
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
 \end{table}

@@

\subsection{Boolean circuits}
As a toy domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates,
as well as 2 or 4 input multiplexers.
This experimental setup is an extension of one introduced in the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
For this domain our recognition model is a simple multilayer perceptron
whose inputs are the input/output examples.
Figure~\ref{circuitLearningCurve} shows our algorithm learning to build circuits; Table~\ref{circuitExperimentTable}

@@

The neural recognition model accelerates inference,
but given enough iterations of our algorithm,
a lesioned version without the recognition model eventually comes close to catching up (red curves in Figure~\ref{regressionLearningCurve}; No NN in Table~\ref{regressionComparison}).
We compare our model on held-out tasks against 4 baselines (Table~\ref{regressionComparison}):
No NN, which lesions $q$;
RF/DC,
which holds the generative model $(\mathcal{D},\theta)$ fixed and learns
a $q$ only from samples from the fixed generative model --
this is how RobustFill and DeepCoder are trained;
PCFG, which lesions $q$, fixes $\mathcal{D}$, and learns only $\theta$ (so it is analogous to learning the parameters of a PCFG);
and Enum, which does no learning and just enumerates a large frontier.
\begin{figure}\centering
  \includegraphics[width = 7cm]{figures/regression.eps}
  \caption{Learning curves for symbolic regression. Green: \system. Red: no neural network. Notice how the model without the neural network eventually almost catches up.}\label{regressionLearningCurve}
\end{figure}

@@

\subsection{String editing}\label{textSection}
Synthesizing programs that manipulate strings is a classic problem in the
programming languages and AI literatures~\cite{menon2013machine,lau2001programming},
and algorithms that learn string editing programs ship in Microsoft Excel~\cite{gulwani2011automating}.
However, this prior work presumes a ready-made DSL,
expertly crafted to suit string editing.
We show in this experiment that we can instead start out with a
more restricted set of primitives -- close to the string routines provided by Python --
and automatically recover many of the higher-level building blocks that have made these
other system successful.
Table~\ref{textPrimitives} lists the primitive functions provided to \system.
Absent are any routines for searching for characters, extracting substrings,
or replacing one character with another.
Instead, we provide array/string indexing primitives (\texttt{slice}, \texttt{nth}),
basic arithmetic, and functions for splitting a string into a list (\texttt{split})
or turning a list into a string (\texttt{join}).

We automatically generated  600 string editing along with input/output examples (Figure~\ref{exampleTextProblem}), split into a 75/25 train/test split.

@@

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}, \texttt{trim}, \texttt{nth}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}  
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\texttt{c}$) = $\texttt{(+ (chr->str c))}\\
      $f_1($\texttt{i,c,s}$) = $      \texttt{(+ (nth i (split c s))}\\\hspace{2.55cm}\texttt{ (chr->str c))}\\
      $f_2($\texttt{s,c}$) = $\texttt{((}$f_0$\texttt{ c) (}$f_1$\texttt{ (1+ 0) c s))} \\
      \hspace{1cm}($f_2$: \emph{Extract substring delimited by character})
    %% $f_0($\texttt{i,c,s}$) = $    \texttt{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\texttt{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}
%% \begin{figure}

@@

\begin{figure}
  \begin{tabular}{ll}\toprule
    Input&Output\\\midrule
    kIDY eMSz,zON& 	eM\\
KCQ njp,yzhq 	&nj\\
zeChf zzZ,dtFP 	&zz\\
plbKi fmLcO,ykkOR 	&fm    \\\bottomrule
  \end{tabular}
  \caption{Example string edit problem (left) and the program that our system recovers (right)}\label{exampleTextProblem}
  \end{figure}

>>>>

How should we perform inference for the model
in~\ref{graphicalModel}(a)?
We take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a wake cycle means using the recognition model to search for programs that solve the tasks.
The wake cycle can be thought of as a kind of active problem solving,
guided by the current generative and recognition models.
A sleep cycle means updating these models so that the DSL and  recognition model
to more closely match the distribution of tasks.
This wake/sleep style of inference is the reason why we call our approach \system.

@@

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions the domain (Table~\ref{initialExampleDSL}).
 \begin{table}
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\texttt{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \texttt{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \texttt{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists& \texttt{(map (lambda (x) (+ x a)) l)}\\
   &\hspace{0.5cm} \emph{(add a to every element of list l)}
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
 \end{table}

@@

\subsection{Boolean circuits}
As a toy domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 250 random circuits made out of
AND, NOT, and OR gates,
as well as 2 or 4 input multiplexers.
This experimental setup is an extension of one introduced in the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
%% For this domain our recognition model is a simple multilayer perceptron 
%% whose inputs are the input/output examples.
\system learns to build these
circuits by inventing new subroutines that correspond to
logic gates and multiplexers (Figure~\ref{circuitTables}, top).
We compare  with four baselines on 250 held-out circuits:
No NN, which lesions $q$;
RF/DC,
which holds the generative model $(\mathcal{D},\theta)$ fixed and learns
a $q$ only from samples from the fixed generative model --
this is how RobustFill and DeepCoder are trained;
PCFG, which lesions $q$, fixes $\mathcal{D}$, and learns only $\theta$ (so it is analogous to learning the parameters of a PCFG);
and Enum, which does no learning and just enumerates a large frontier.
\begin{figure}
\begin{tabular}{lllll}
  \toprule \system&No NN  & RF/DC & PCFG & Enum\\\midrule
  84\% solved&76\%& \\
  $0.9$ sec&  1.1
  \\\bottomrule
\end{tabular}
\end{figure}

@@

The neural recognition model accelerates inference,
but given enough iterations of our algorithm,
a lesioned version without the recognition model eventually comes close to catching up (green curves in Figure~\ref{regressionLearningCurve}; No NN in Table~\ref{regressionComparison}).
We compare our model on held-out tasks against 4 baselines (Table~\ref{regressionComparison}).
\begin{figure}\centering
  \includegraphics[width = 7cm]{figures/regression.eps}
  \caption{Learning curves for symbolic regression. Red: \system. Green: no neural network. Notice how the model without the neural network eventually solves almost all the tasks, but does so using programs with lower likelihood (see Table~\ref{regressionComparison}).}\label{regressionLearningCurve}
\end{figure}

@@

\subsection{String editing}\label{textSection}
Synthesizing programs that manipulate strings is a classic problem in the
programming languages and AI literatures~\cite{menon2013machine,lau2001programming},
and algorithms that learn string editing programs ship in Microsoft Excel~\cite{gulwani2011automating}.
However, this prior work presumes a ready-made DSL,
expertly crafted to suit string editing.
We show in this experiment that we can instead start out with a
less specialized set of primitives -- close to the string routines provided by Python --
and automatically recover many of the higher-level building blocks that have made these
other system successful.
Table~\ref{textPrimitives} lists the primitive functions provided to \system.
Absent are any routines for searching for characters, extracting substrings,
or replacing one character with another.
Instead, we provide array/string indexing primitives (\texttt{slice}, \texttt{nth}),
basic arithmetic, and functions for splitting a string into a list (\texttt{split})
or turning a list into a string (\texttt{join}).

We automatically generated  600 string editing along with input/output examples (Figure~\ref{exampleTextProblem}), split  75/25 train/test.

@@

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}, \texttt{trim}, \texttt{nth}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}  
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\texttt{c}$) = $\texttt{(+ (chr->str c))}\\
      $f_1($\texttt{i,c,s}$) = $      \texttt{(+ (nth i (split c s))}\\\hspace{2.55cm}\texttt{ (chr->str c))}\\
      $f_2($\texttt{s,c}$) = $\texttt{((}$f_0$\texttt{ c) (}$f_1$\texttt{ (1+ 0) c s))} \\
      \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\texttt{c,s}$) = $\texttt{(nth (1+ 0) (split c (nth 0 s)))}\\
      \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\texttt{i,c,s}$) = $    \texttt{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\texttt{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}
%% \begin{figure}

@@

\begin{figure}
  \centering\begin{tabular}{ll}\toprule
    Input&Output\\\midrule
cqbHT$>$aIKd$<$szW& 	AIKD\\
YBW$>$TIg$<$WSho 	&TIG\\
DkI$>$DwKI$<$xaPoc& 	DWKI\\
KfvTh$>$Dbl$<$OIJso 	&DBL
    \\\bottomrule
  \end{tabular}

  \vspace{0.5cm}

  \begin{minipage}{7.5cm}
 $f($\texttt{s}$) = $   \texttt{(}$f_3$\texttt{ (split '>' (upper s)) '<')}
    \end{minipage}
  \caption{Example string edit task (top) and program  \system writes (bottom). $f_3$ is a new DSL subroutine discovered by \system, defined in Table~\ref{textPrimitives}}\label{exampleTextProblem}
  \end{figure}
